# Introduction {.unnumbered}

## How to install R and set up environment

Refer to <https://frbcesab.github.io/rsetup/>


### R packages needed for these tutorials

We will mostly use `terra` and `mapview` as a starter.
Fancier maps can be created with mapsf and tmap (*ref tutorials*)

```{r}
#| label: install packages
needed_packages <- c("terra", "sp", "mapview", "mapsf", "tmap")

check_and_install <- function(x) {
  if (!requireNamespace(x, quietly = TRUE)) {
    install.packages(x)
  }
}

invisible(lapply(needed_packages, check_and_install))
```

## Download needed dataset

As an toy dataset, the tutorials rely on observed occurrence of otter in 2021 around Montpellier, France. These dataset were downloaded directly from R, see the [tutorial on creating a toy dataset from GBIF and IGN](https://frbcesab.github.io/spatial-r/chapters/31_toydata_mpl.html) for more details.

You can run the following command to download the dataset (total size : 9Mb).
```{r}
#| eval: false
# set data directory
datadir <- here::here("data")
if (!dir.exists(datadir)) {
  dir.create(
    path = datadir,
    showWarnings = FALSE,
    recursive = TRUE
  )
}

# download base url
url_git <- "https://github.com/FRBCesab/spatial-r/raw/refs/heads/main/data"

# download GBIF data (20kb)
file1 <- "gbif_otter_2021_mpl50km.csv"
download.file(file.path(url_git, file1), file.path(datadir, file1), mode = "wb")

# download elevation data (500kb)
file2 <- "BDALTI_mpl50km.tif"
download.file(file.path(url_git, file2), file.path(datadir, file2), mode = "wb")

# download administrative borders (2.5Mb)
file3 <- "BDCARTO-Commune_mpl50km.gpkg"
download.file(file.path(url_git, file3), file.path(datadir, file3), mode = "wb")

# download rivers (2Mb)
file4 <- "BDCARTO-River_mpl50km.gpkg"
download.file(file.path(url_git, file4), file.path(datadir, file4), mode = "wb")

# download land cover (3.7Mb)
file5 <- "BDCARTO-LULC_mpl50km"
shpext <- c("dbf", "prj", "shp", "shx")
url_shp <- file.path(url_git, paste(file5, shpext, sep = "."))
dir_shp <- file.path(datadir, paste(file5, shpext, sep = "."))
for (i in seq_along(shpext)) {
  download.file(url_shp[i], dir_shp[i], mode = "wb")
}
```

