# Grid {.unnumbered}

:::{.callout-important title="Summary"}
This tutorial explore how to make a grid and transform observations into a grid with `terra` package:

- import a multi-layer rasters with [`terra::rast()`](https://search.r-project.org/CRAN/refmans/terra/html/rast.html)   
- extract information from rasters on points or polygons with [`terra::extract()`](https://search.r-project.org/CRAN/refmans/terra/html/extract.html)

:::

:::{.callout-tip title="The ecologist mind"}
If we want to make some modeling, it can be needed to regroup observations into a grid.
:::


## Setup

<details>
<summary> Follow the setup instructions if you haven't followed previous tutorials</summary>

If haven't done it already, please follow the [setup instructions](https://frbcesab.github.io/spatial-r/chapters/01_setup.html).    

Let's start with loading the required packages.

```{r setting}
suppressPackageStartupMessages({
  library(mapview)
  library(here)
  library(terra)
  # library(sf)
  # library(exactextractr)
})
```

Now load all the datasets attached to the observations of otters recorded in 2021 within a 50km buffer from Montpellier, France.  

```{r data_local}
pt_otter <- vect(here("data", "gbif_otter_2021_mpl50km.gpkg"))
river <- vect(here("data", "BDCARTO-River_mpl50km.gpkg"))
landuse <- vect(here("data", "BDCARTO-LULC_mpl50km.shp"))
bdalti <- rast(here("data", "BDALTI_mpl50km.tif"))
temperature <- rast(here("data", "CHELSA_monthly_tas_2015_2021.tif"))
```

</details>

## Create a grid

Key issue : which resolution? which projection?
resolution depends on your dataset, extent, and ecological question
projection depends on the dataset

Let's start from our observation.
```{r make_grid}
pt_2154 <- project(pt_otter, "EPSG:2154")
res <- 5000 # 1km grid
grid <- rast(pt_2154, res = res)
grid
```

## Points to grid

the argument `fun=count` set that it calculate the number of points
the argument `background=0` set that f there is no point, the vale will be 0, instead of NA by default
```{r grid_nobs}
grid_nobs <- rasterize(pt_2154, grid, fun = "count", background = 0)

plot(grid_nobs)
```

## Lines to grid   

length of rivers

```{r grid_river}
river_2154 <- project(river, "EPSG:2154")

grid_river <- rasterizeGeom(river_2154, grid, fun = "length")
plot(grid_river)
```


## Polygon to grid

percentage of forest

```{r grid_forest}
# select only the forest
forest <- landuse[landuse$nature == "Forêt"]
# project the polygons
forest_2154 <- project(forest, "EPSG:2154")
# rasterize
grid_forest <- rasterize(forest_2154, grid, cover = TRUE, background = 0)

plot(grid_forest)
```

## Resample raster

### Elevation

```{r grid_alti}
grid_alti <- resample(bdalti, grid, "bilinear")
plot(grid_alti)
```

### Temperature

```{r grid_temp}
#calculate the average temperature in 2021
in_2021 <- grep("2021", names(temperature))
monthly_temp <- subset(temperature, in_2021)
annual_temp <- mean(monthly_temp, na.rm = TRUE)
# project (that's one of the only occasion when projection of raster)
temp_2154 <- project(annual_temp, "EPSG:2154")
grid_temp <- resample(temp_2154, grid, "bilinear")
plot(grid_temp)
```

## Merge all together

Ignoring all concept of spatial bias or anything else
```{r merge}
out <- c(grid_nobs, grid_river, grid_forest, grid_alti, grid_temp)

names(out) <- c("nobs", "river_m", "forest", "elevation_m", "temperature_K")

plot(out)
```

```{r mask}
france_border <- geodata::gadm("FRA", level = 0, path = here("data"))
# project the coast in Lambert-93
france_2154 <- project(france_border, crs(bdalti))
# mask (=set to NA) the pixels that are not in the polygon
out_masked <- mask(out, france_2154)
plot(out_masked)
```

```{r anova}
df <- data.frame(out_masked)

m1 <- lm(nobs ~ river_m + forest + elevation_m + temperature_K, data = df)
anova(m1)
```

:::{.callout-note title="Your turn"}
Does the spatial resolution has en effect on the results? Repeat the same process with a grid of 500m, 1km, or 10km resolution and compare the results. 
:::
